---
title: "R Notebook"
output:
  html_notebook: default
  html_document:
    df_print: paged
---
data = filter(data, (data$winner=="white" & data$white_rating > data$black_rating) | (data$winner=="black" & data$white_rating < data$black_rating))

```{r}
library(ggplot2)
library(plyr)
library(corrplot)
library(stringr)


data = read.csv("games.csv")
data$bg_winner = (data$winner=="white" & data$white_rating > 1.2*data$black_rating) | (data$winner=="black" & 1.2*data$white_rating < data$black_rating)
data$diff = abs(data$white_rating - data$black_rating)
data$winner = data$winner %in% c("white", "draw")
data$rt_bg = data$white_rating > 1.4*data$black_rating | 1.4*data$white_rating <data$black_rating

mean(data$rt_bg)


data1 = data[c("turns", "bg_winner", "diff", "rt_bg")]
cor = rquery.cormat(data1)
cor
ggplot(data, aes(x=turns, y=..density.., fill=rt_bg)) + geom_density(alpha=.5)
#data$turns = as.integer(data$turns)
#data$rate_diff = data$white_rating - data$black_rating
#means = ddply(data, "opening_name", summarise, grp.mean = mean(turns))
#means
#ggplot(data, aes(x=rate_diff, fill=winner)) + geom_density(alpha=0.5)
```
# Intro

As the topic of the research I, as a big fan, chose chess. The dataset I used in 
the analysis was found in kaggle (https://www.kaggle.com/datasnaek/chess). It
contains a set with over 20.000 chess games played in 2016-2018 at one of 
the most popular online chess platforms Lichess(https://lichess.org/). Each game
includes the id of the players, the winner, the type of victory (checkmate, 
resigning, exceeded time limit or draw), number of turns, the ratings of the
players, info about the opening and other cool stuff.

```{r}
library(ggplot2)
library(plyr)
library(corrplot)
library(stringr)
library(fitdistrplus)
data = read.csv("games.csv")
head(data)
```

# Aim
Obviously, one the most outstanding thing we can do with this dataset is to
predict the winner of the game. In order to do that, let's decide what info in 
the dataframe can be used for such prediction. Let's have a look at the
correlation matrix constructed of relevant columns of dataframe.

```{r}
# Clear the data, we are not interested in game played in draw
data = data[which(data$winner != "draw"), ]
data$time = as.integer(str_extract(data$increment_code, "\\d+")) # increment code
# is a convention of the given time for each player to make moves, by extracting
# the first part we get time for moves (second part is an additional time in 
# seconds after each move is made)
data$is_white_winner = data$winner == "white" # need boolean values for creating
#corr matrix
data$win_through_mate = data$victory_status == "mate"

data$rated = data$rated == "TRUE"
data = data[c("rated", "turns", "white_rating", "black_rating", "is_white_winner",
              "time", "win_through_mate")]
matrix = rquery.cormat(data)
```

Unfortunately, correlation matrix doesn't show any significant dependency between
the winner and  some other features that could be used later in modeling
logistic regression. Therefore, did we come to the dead end? Actually, there is 
a thing we could try to do else. The ratings of white player and black separately
doesn't actually help our prediction. However, the difference between the player's
ratings seems like a logical thing to suggest, when trying to predict the winner
of the outcome of the game. Let's replace rating of white and black players with
their difference and see how the correlation matrix changes.

```{r}
data$rating_diff = data$white_rating - data$black_rating # note that we don't 
# take absolute value of difference
data = data[c("rated", "turns", "is_white_winner", "time", "win_through_mate",
              "rating_diff")]
matrix = rquery.cormat(data)
```

That's much better, the difference in ratings and the winner are indeed dependent
Before jumping straight to modeling our logistic regression, let's analyze the
distribution of the difference between ratings, as it's our key to determining the
winner and it is actually interesting to see how players are being selected at 
this online platform. My first suggestion is that the distribution is normal, as
this selection should be carefully selected usually with same rating with 
normal deviation (so it would be much more interest in game for both players)

```{r}
ggplot(data, aes(x=rating_diff)) +
  geom_histogram(aes(y=..density..), color="black", fill="white", binwidth = 80) + 
  geom_density(alpha=.5, fill="red") + 
  stat_function(fun=dnorm, geom="area", args=list(mean=mean(data$rating_diff), sd=sd(data$rating_diff)), color="blue", alpha=.3) +
  labs(x="Difference between white player and black player ratings", y="Density", 
       title="Distribution of rating difference")
```
Having a look at the density of our distribution, we can conclude that it really 
resembles normal one quite a lot, however it is much peakier than the normal.
First of all, that should mean that the kurtosis of our distribution is much
higher than in normal one (3), while skewness should be near 0 (because 
it is almost perfectly symmetrical). Take a look at the summary, to see it.
```{r}
descdist(data$rating_diff)
```
Summary proves our assumptions made on the basis of graphic. Cullen and Frey 
graph also suggests that our distribution is closer more to logistic 
than to a normal one. Therefore, I just googled logistic distribution and indeed: "It 
resembles the normal distribution in shape but has heavier tails (higher kurtosis)." 
Considering all said above, our hypothesis is that the distribution of rating
difference is logistic. Before making any hasty decision, let's conduct a more
detailed visual analysis to see if we can relate our distribution to a logistic one.

```{r}
fit.logistic = fitdist(data$rating_diff, "logis")
#fit.logistic
#plot(fit.logistic)
fit.norm = fitdist(data$rating_diff, "norm")
fit.norm
fit.norm$estimate
```

Okay, the empirical cdf looks like converging to the theoretical one, Q-Q plot
compares quantiles of our distribution and the theoretical one - they follow almost
a straight line, which means that there is almost no deviation of our distribution
from the theoretical logistic one. Similarly with P-P plot. Considering all of this, 
can we finally conclude that we are dealing with logistic distribution? Visualization techniques are always good, but, unfortunately, we can't say anything for sure without
a formal statistical test. Let's carry out one! For testing the distribution family, 
we use a Kolgomorov-Smirnov test. By LLN, the empirical cdf should converge to the
theoratical one, as sample size increases. By Glivenko–Cantelli theorem, the maximum 
difference between respective points of ecdf and cdf -> 0 with probability 1.
$\sup_{t\in R}|\hat{F_{x, n}}(t) -  F_x(t)| -> 0$. Therefore, if this difference is
small enough, we say that our distribution belongs to the family we assumed. That is
the logic behind Kolgomorov Smirnov test.

```{r}
ks.test(data$rating_diff, "plogis", fit.logistic$estimate[1], fit.logistic$estimate[2])
ks.test(data$rating_diff, "pnorm", fit.norm$estimate[1], fit.norm$estimate[2])
```